---
title: "Exploring patterns of environmental justice"
author: "Josephine Cardelle"
format: html
editor_options: 
  chunk_output_type: console
---

### Load packages

```{r, echo = T, results = 'hide'}
# Load libraries
library(tidyverse)
library(here)
library(tmap)
library(sf)
library(dplyr)
library(kableExtra)
library(testthat)
```

## Part 1: Legacy of redlining in current environmental (in)justice

a map of historical redlining neighborhoods, including:

-   neighborhoods colored by HOLC grade
-   an appropriate base map
-   a table summarizing the percent of current census block groups within each HOLC grade (or none)
-   a set of figures summarizing current conditions (from the EJScreen data) within HOLC grades using the mean of the following variables:
    -   \% low income
    -   percentile for Particulate Matter 2.5
    -   percentile for low life expectancy
    -   a brief paragraph reflecting on these result

### Import Data

```{r, echo = T, results = 'hide'}
# Read in data
ej <- st_read(here("data/ejscreen/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))
la_red <- st_read(here("data/mapping-inequality/mapping-inequality-los-angeles.json")) %>%
st_make_valid()
```

```{r}
# Filter ej data so ID in water is not showing
ej <- ej %>%
  filter(ID != "060379903000" & ID != "060379902000")
# Filter ej data to just LA
la_ej <- ej %>%
filter(CNTY_NAME %in% c("Los Angeles County")) 

# Warning and transformation if coordinate reference systems do not match
if(st_crs(la_ej) != st_crs(la_red)){
  warning("updating coordinate reference systems to match")
  la_red <- st_transform(la_red, crs = st_crs(la_ej))
}

```

### Visualize Data

```{r}

# Map making
tm_shape(ej, bbox = la_red) + # Base map and set bounding box to redline data
  tm_polygons(col = "#FFFEEB",
              border.col = "black",
              lwd = 0.3) +
tm_shape(la_red) +
tm_fill("grade",
        title = "HOLC grade",
        palette = c("#FD151B", "#FFB30F", "#849324", "#437F97")) +
  tm_scale_bar(position = c("right", "top")) +
  tm_compass(position = c("right", "top"),
             size = 1.5,
             text.size = 0.8) +
  tm_layout(main.title = "Home Owners’ Loan Corporation Grades in LA County",
            main.title.position = c("left", "top"),
           legend.outside = TRUE,
           bg.color = "lightblue")



  
# Census block groups within all HOLC
# Join using st_intersects which encompasses all the spatial intersect relationships, and used left=FALSE for an inner join. An inner join will return records that spatially match
red_cbg <- st_join(la_ej, la_red, join = st_intersects, left = FALSE)

# Create table with percentages for each HOLC grade
percent_holc <- red_cbg %>%
group_by(grade) %>%
summarise(count = n()) %>%
mutate(percent = (count / sum(count))*100 ) %>% 
select(-count) %>% 
st_drop_geometry()


# Use testthat package to make sure all percents add to 100
expect_equal(sum(percent_holc$percent), 100)

# Create table
kable(percent_holc, caption = "Percent of current census block groups within each HOLC grade")

# Table with means for 3 given variables
holc_ej_means <- red_cbg %>%
group_by(grade) %>%
summarise(lowincpct = mean(LOWINCPCT, na.rm = TRUE),
pm25 = mean(P_PM25, na.rm = TRUE),
lifeexppct = mean(P_LIFEEXPPCT, na.rm = TRUE))


# Make figures summarizing variables in each HOLC grade
ggplot(holc_ej_means, aes(x = grade, 
                          y = lowincpct, 
                          fill = grade)) +
  geom_col() +
  scale_fill_manual(values = c("#FD151B", "#FFB30F", "#849324", "#437F97")) +
  labs(x = "HOLC Grade", 
       y = "Percent Low Income", 
       fill = 'HOLC Grade') +
  ggtitle("HOLC Grade and % Low Income in LA County")

ggplot(holc_ej_means, aes(x = grade, 
                          y = pm25, 
                          fill = grade)) +
  geom_col() +
  scale_fill_manual(values = c("#FD151B", "#FFB30F", "#849324", "#437F97")) +
  labs(x = "HOLC Grade", 
       y = "Percentile for Particulate Matter 2.5", 
       fill = 'HOLC Grade') +
  ggtitle("HOLC Grade and Percentile for PM2.5 in LA County")

ggplot(holc_ej_means, aes(x = grade, 
                          y = lifeexppct, 
                          fill = grade)) +
  geom_col() +
  scale_fill_manual(values = c("#FD151B", "#FFB30F", "#849324", "#437F97")) +
  labs(x = "HOLC Grade", 
       y = "Percentile for Low Life Expectancy", 
       fill = 'HOLC Grade') +
  ggtitle("HOLC Grade and Percentile for Low Life Expectancy in LA County")

```

The results showed the highest values for % low income, percentile for PM 2.5, and percentile for low life expectancy in HOLC grade D. The lowest for all three variables was shown in HOLC grade A. This shows extreme inequalities in the different grades in economy, environment, and health.

## Part 2: Legacy of redlining in biodiversity observations

For this assignment, you must produce the following based on observations from 2022:

-   a figure summarizing the percent of observations within redlined neighborhoods within each HOLC grade

-   a brief paragraph explaining whether these results match the findings from Ellis-Soto et al. 2023

```{r}
# Read in bird data
birds <- read_sf(here("data","gbif-birds-LA", "gbif-birds-LA.shp"))

# Filter to only 2022
birds_22 <- birds %>% filter(year == 2022)

```

```{r}
# Warning and transformation if coordinate reference systems do not match
if(st_crs(birds_22) != st_crs(la_ej)){
  warning("coordinate refrence systems do not match")
  birds_22 <- st_transform(birds_22, crs = st_crs(la_ej))
}
```

```{r}
# Join using st_intersects which encompasses all the spatial intersect relationships, and used left=FALSE for an inner join. An inner join will return records that spatially match
# Drop NAs so that only the percentage of observations within each HOLC grade, relative to HOLC grades only are are shown
red_birds <- st_join(x = la_red, y = birds_22, join = st_intersects, left = FALSE) %>% 
  filter(!is.na(grade))

# Create table of percent of bird observations in HOLC grades
birds_percent <- red_birds %>%
group_by(grade) %>%
summarise(count = n()) %>%
mutate(percent = (count / sum(count))*100 ) %>% 
select(-count) %>% 
st_drop_geometry()

# Return table
kable(birds_percent, caption = "Percent of bird observations in redlined neighborhoods in each HOLC grade")


# Create plot of bird observations in each HOLC grade
ggplot(birds_percent) +
geom_col(aes(x = grade, 
             y = percent, 
             fill = grade)) +
scale_fill_manual(values = c("#FD151B", "#FFB30F", "#849324", "#437F97")) +
labs(x = "HOLC grade", 
     y = "Percentage of observations", 
     fill = 'HOLC Grade') +
  ggtitle("Bird Observations in Redlined Neighborhoods in 2022")
```

These results do not match with the finding from Ellis-Soto et al. 2023. The article says that areas previously redlined are often underrepresented in biodiversity data collection. However, the results from this table and graph show that the highest percent of bird observations within redlined neighborhoods in LA County were seen in grade C neighborhoods, followed by grade D, then grade B and A, respectively. However, this could be that there are more grade C and D areas in LA County than A and B. To do a real analysis we would have to take into account the area of each HOLC grade in LA County.

## Data Citations

| Data                                                       | Citation                                                                                                                                                                                                                                                      | Link                                                                                                                                                                |
|------------------------------------------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| HOLC Redlining                                             | **Digital Scholarship Lab.** *Redlining Map Data (JSON file).* University of Richmond, Digital Scholarship Lab. Accessed October 17, 2024.                                                                                                                    | <https://dsl.richmond.edu/panorama/redlining/data>                                                                                                                  |
| Biodiversity Observations                                  | Data from: the Global Biodiversity Information Facility                                                                                                                                                                                                       | <https://www.gbif.org/>                                                                                                                                             |
| EJScreen: Environmental Justice Screening and Mapping Tool | United States Environmental Protection Agency. 2023. Data from: Geodatabase of National EJScreen Data at the Block Group Level                                                                                                                                | [Geodatabase of National EJScreen Data at the Block Group Level](https://gaftp.epa.gov/EJScreen/2024/2.31_August_useMe/EJScreen_2024_BG_with_AS_CNMI_GU_VI.gdb.zip) |
| Historical redlining and biodiversity sampling artile      | Ellis-Soto, D., Chapman, M. & Locke, D.H. Historical redlining is associated with increasing geographical disparities in bird biodiversity sampling in the United States. *Nat Hum Behav* **7**, 1869–1877 (2023). https://doi.org/10.1038/s41562-023-01688-5 | <https://www.nature.com/articles/s41562-023-01688-5#citeas>                                                                                                         |
